Resource Adequacy and Operational Security Interaction in the EPOC 2030-50 Project

### 2022-11-30 papers/main.tex

\begin{abstract}
    This report outlines an attempt to integrate a resource adequacy and operational security exercise. The idea is to use a unit commitment schedule which is then used as the input to a more detailed power flow model. An additional goal was to make explicit the trade-off between day ahead adequacy and real time reliability. We used an adapted version of KU Leuven's \ac{DUCPR} model which accounts for network constraints in the reserve activation phase for this purpose. However the \ac{DUCPR} model is not able to capture enough detail in the reserve activation phase for this tradeoff to be made. This and other omissions of the \ac{DUCPR} model when compared to the power flow model meant that the load shedding observed in the two models differed by an order of magntiude. All code for the adequacy side of this exercise can be found at \href{https://github.com/junglegobs/ASoSEPOC}{https://github.com/junglegobs/ASoSEPOC}.
\end{abstract}

@@

As noted in the abstract, we were unsuccesful in illustrating the value of this interaction for two reasons. Firstly, we were unable to illustrate the tradeoff between adequacy and security with only the \ac{DUCPR} model due to its approximate treatment of real time network imbalances due to reserve activation. Secondly and related, the expected load shedding resulting from the \ac{DUCPR} differed by an order of magnitude to that found by the power flow, likely due to due to the difference in the treatment of storage as well as the more detailed network representation in the power flow model.

@@

The rest of this document is organised as follows. \secref{sec:model} describes the \ac{DUCPR} model used to simulate power system operations and obtain day ahead adequacy and very approximate real time operational security indicators, namely scheduled load shedding and reserve shedding. \secref{sec:data} describes the stylised Belgian system analysed. \secref{sec:initial_results} analyses a single day in order to identify sources of adequacy and security issues, while \secref{sec:adequacy_vs_security} compares the \ac{DUCPR} model results to that of the power flow analysis.

@@

It is possible to calculate $d_{nt}^{min}$ and $d_{nt}^{max}$ since the quantiles for $D^{L^+}$ and $D^{L^-}$ are calculated based on forecast error scenarios which \emph{are} defined at the nodal level. These limits are shown in Figure \ref{fig:imbalance_range}. Clearly it is simply not possible to have an imbalance on some nodes, such as Coo (the location of Belgium's pumped hydro unit). Constraining $d_{nlt}^{L^+}$ and $d_{nlt}^{L^-}$ in this way is referred to as AbsImb later on.

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.9/textwidth]{../plots/process_input/residual_load_timeseries}
    \caption{Residual load time series of test system.\label{fig:residual_load_ts}}
\end{figure}

@@

@@

This analysis relates only to the 285th day of the year (12th of October). The aim is to investigate what are the causes of load shedding to better understand the system.

@@

\tabref{tab:results_no_OR} lists results for an increasing level of technical constraints or increasingly inflexible system. While the objective increases by \euro 1,267,000 from the simple linear economic dispatch model with no network constraints to the unit commitment model with network constraints, no load shedding occurs. Preventing simultaneous charging and discharging constraints actually decreases, though this difference is within the optimality gap (0.01\%). This suggests that additional energy consumption (which is not physically possible) from storage does not aid congestion in this context.

@@

\tabref{tab:results_no_RANet} summarises results for the model runs with probabilistic operating reserves but no reserve activation network constraints. Clearly unit commitment constraints lead to greatly increased load shedding, 3.8 GWh to be precise. They also lead to reserve shedding where there wasn't any before. Neither day ahead network constraints or preventing simultaneous charging or discharging have an effect on the objective function which is greater than the MIP gap.

@@

\figref{fig:dispatch_with_operating_reserves} shows the dispatch schedule for the case with unit commitment constraints, day ahead network constraints and operating reserve requirements but no limit on reserve shedding. Load shedding occurs at 8 - 10 AM, even though there is still energy in the storage units. This is not due to congestion issues, since some of these storage units are located at the same node as the load being shed. Rather, this is likely due to the constraint which specifies that the storage state of charge at the end of the day must be greater than or equal to that at the start of the day. This constraint was included so as not to overestimate the flexibility that storage could provide, though it is unlikely that such a situation would occur in reality.

@@

Considering the case with just unit commitment constraints, the reserve shedding limit can be decreased without increasing load shedding, though there is an increase (~0.09\%) in the objective function greater than the MIP gap. This is counterintuitive, as you would expect the DUC-PR model to shed all operating reserves before shedding load in the day ahead stage. The former is less costly than the latter after all, due to activation probabilities being < 1. However, what is occuring is most likely `economic reserve shedding' - shedding reserves because they are highly unlikely to be activated. This phenomenon was also observed in~\cite{Hermans2018}. \figref{fig:reserves_with_operating_reserves} illustrates that almost all reserve shedding occurs between 14:00 and 16:00 (hour 6830 to 6832) and not 8:00 to 10:00 as with load shedding. These are hours with a lot of solar generation and very little conventional, so to avoid this reserve shedding only requires additional generator commitment and not increasing the amount of load shedding in those hours.

@@

Multiplying the load by a factor of 1.5 gives the results shown in \tabref{tab:results_with_RANet_LM_15}. Here it is possible to discern a trade-off between load shedding and reserve shedding, albeit a small one - reducing reserve shedding to 0 from approximately 12,000 MWh increases load shedding only by 914 MWh. That this effect is slight is for similar reasons to that listed previously - reserve shedding occurs in different hours to load shedding and load shedding occurs in the same hours as before (8:00 to 11:00) and is likely due to the cyclic constraint on storage.

@@

\section{Day ahead adequacy and real time operational security trade offs} \label{sec:adequacy_vs_security}

\subsection{Simple illustrative case}

\begin{figure}[h]
    \centering
    \begin{subfigure}[t]{0.9\textwidth}
        \centering
        \includegraphics[width=\textwidth]{../plots/copperplate_adequacy_security_tradeoff/load_shedding_vs_reserve_shedding}
        \caption{Reserve shedding plotted as a function of load shedding}
    \end{subfigure}
    \begin{subfigure}[t]{0.9\textwidth}
        \centering
        \includegraphics[width=\textwidth]{../plots/copperplate_adequacy_security_tradeoff/load_shedding_vs_expected_load_shedding}
        \caption{Expected additional real time load shedding (reserve shedding weighted by activation probability) plotted as a function of load shedding.}
    \end{subfigure}
\caption{Simple illustration of the potential trade-off between planned load shedding (i.e. day ahead adequacy) and expected load shedding (i.e. real time operational security). Results are for day 161, no storage, unit commitment or network constraints are included.\label{fig:simple_reserve_vs_load_shedding}}
\end{figure}

\figref{fig:simple_reserve_vs_load_shedding} is included here to illustrate the trade-off we wanted to be able to investigate. However, for the reasons described in \secref{sec:initial_results} and \bref{sec:ULiege_comparison}, this was not possible.

\subsection{Comparison with ULiege operational security results} \label{sec:ULiege_comparison}

A comparison of the real time operational security as percieved by the DUC-PR model compared to those obtained from a power flow performed by the team ULiege is given in the figure below.

\begin{figure}[h]
    \centering
    \begin{subfigure}[t]{0.9\textwidth}
        \centering
        \includegraphics[width=\textwidth]{../plots/compare_ULiege_results/forecast_only_no_contingency.pdf}
        \caption{Day ahead load shedding for DUC-PR and operational security power flow model.}
    \end{subfigure}
    \begin{subfigure}[t]{0.9\textwidth}
        \centering
        \includegraphics[width=\textwidth]{../plots/compare_ULiege_results/with_scenarios_no_contingency.pdf}
        \caption{Expected load shedding for DUC-PR and operational security power flow model.}
    \end{subfigure}
\caption{Results are for day 309, with storage, unit commitment and network (reserve activation) constraints.\label{fig:Uliege_compare}}
\end{figure}

Clearly the two models disagree. We suggest the following non-exhaustive list of reasons why this could be:

\begin{itemize}
    \item The approximate modeling of real time imbalances in the DUC-PR model (see \secref{sec:operating_reserve_network_activation_constraints} and \secref{sec:further_constraining_operating_reserve_network_activation_constraints}). As extensively discussed, this means the DUC-PR model overestimates the ease with which real time imbalances may be dealt with.
    \item The omission of N-1 contingencies in the DUC-PR model. This omission also leads to an overestimation of the ease of dealing with real time imbalances.
    \item The extra-liberty for storage to deal with real time imbalances in ULiege's power flow model. This has the opposite effect.
\end{itemize}

>>>>

\begin{abstract}
    This report outlines an attempt to integrate a resource adequacy and operational security exercise. The idea is to use a unit commitment schedule which is then used as the input to a more detailed power flow model. An additional goal was to make explicit the trade-off between day ahead adequacy and real time reliability. We used an adapted version of KU Leuven's \ac{DUCPR} model which accounts for network constraints in the reserve activation phase for this purpose. However, the presence of storage and the lack of detail captured by the the \ac{DUCPR} model in the reserve activation phase made it difficult to illustrate this tradeoff. This and other omissions of the \ac{DUCPR} model when compared to the power flow model meant that the load shedding observed in the two models differed by an order of magntiude. All code for the adequacy aspect of this exercise can be found at \href{https://github.com/junglegobs/ASoSEPOC}{https://github.com/junglegobs/ASoSEPOC}.
\end{abstract}

@@

As noted in the abstract, we were unsuccesful in illustrating the value of this interaction for two reasons. Firstly, we were unable to illustrate the tradeoff between adequacy and security with only the \ac{DUCPR} model. This was partly because load shedding occurred at different moments to reserve shedding and possibly also due to the approximate treatment of real time network imbalances due to reserve activation in the \ac{DUCPR} model. Secondly and related, the expected load shedding resulting from the \ac{DUCPR} differed by an order of magnitude to that found by the power flow, likely due to due to the difference in the treatment of storage as well as the more detailed network representation in the power flow model.

@@

The rest of this document is organised as follows. \secref{sec:model} describes the \ac{DUCPR} model used to simulate power system operations and obtain day ahead adequacy and very approximate real time operational security indicators, namely scheduled load shedding and reserve shedding. \secref{sec:data} describes the stylised Belgian system analysed. \secref{sec:initial_results} analyses a single day in order to identify the sources of adequacy and security issues, while \secref{sec:adequacy_vs_security} compares the \ac{DUCPR} model results to that of the power flow analysis.

@@

It is possible to calculate $d_{nt}^{min}$ and $d_{nt}^{max}$ since the quantiles for $D^{L^+}$ and $D^{L^-}$ are calculated based on forecast error scenarios which \emph{are} defined at the nodal level. These limits are shown in Figure \ref{fig:imbalance_range}. Clearly it is simply not possible to have an imbalance on some nodes, such as Coo (the location of Belgium's pumped hydro unit). Constraining $d_{nlt}^{L^+}$ and $d_{nlt}^{L^-}$ in this way is referred to as `AbsImb' later on.

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.9\textwidth]{../plots/process_input/residual_load_timeseries}
    \caption{Residual load time series of test system.\label{fig:residual_load_ts}}
\end{figure}

@@

\section{Day ahead adequacy and real time operational security trade offs - a simple illustrative case} \label{sec:adequacy_vs_security}

\begin{figure}[ht]
    \centering
    \begin{subfigure}[t]{0.7\textwidth}
        \centering
        \includegraphics[width=\textwidth]{../plots/copperplate_adequacy_security_tradeoff/load_shedding_vs_reserve_shedding}
        \caption{Reserve shedding plotted as a function of load shedding}
    \end{subfigure}
    \begin{subfigure}[t]{0.7\textwidth}
        \centering
        \includegraphics[width=\textwidth]{../plots/copperplate_adequacy_security_tradeoff/load_shedding_vs_expected_load_shedding}
        \caption{Expected additional real time load shedding (reserve shedding weighted by activation probability) plotted as a function of load shedding.}
    \end{subfigure}
\caption{Simple illustration of the potential trade-off between planned load shedding (i.e. day ahead adequacy) and expected load shedding (i.e. real time operational security). Results are for day 161, no storage, unit commitment or network constraints are included.\label{fig:simple_reserve_vs_load_shedding}}
\end{figure}

\figref{fig:simple_reserve_vs_load_shedding} is included here to illustrate the trade-off we wanted to be able to investigate. However, for the reasons we will describe in \secref{sec:initial_results} and \ref{sec:ULiege_comparison}, this was not possible.

@@

This analysis relates only to the 285th day of the year (12th of October). The aim is to investigate what are the causes of load shedding to better understand the system. We will see that for this particular day, load shedding occurred due to constraint \ref{eq:storage_cyclic_constrant}. This meant that there was no trade-off to be made between load and reserve shedding, unless the load was multiplied by a factor of 1.5. The reserve network activation constraints also had a limited impact on the results.

@@

\tabref{tab:results_no_OR} lists results for an increasing level of technical constraints or increasingly inflexible system. While the objective increases by \euro 1,267,000 from the simple linear economic dispatch model with no network constraints to the unit commitment model with network constraints, no load shedding occurs. Preventing simultaneous charging and discharging constraints actually decreases the objective function, though this difference is within the optimality gap (0.01\%). This suggests that additional energy consumption (which is not physically possible) from storage does not aid congestion in this context.

@@

\tabref{tab:results_no_RANet} summarises results for the model runs with probabilistic operating reserves but no reserve activation network constraints. Clearly unit commitment constraints lead to greatly increased load shedding, 3.8 GWh to be precise. They also lead to reserve shedding where there wasn't any before. Neither day ahead network constraints nor preventing simultaneous charging or discharging have an effect on the objective function greater than the MIP gap.

@@

\figref{fig:dispatch_with_operating_reserves} shows the dispatch schedule for the case with unit commitment constraints, day ahead network constraints and operating reserve requirements but no limit on reserve shedding. Load shedding occurs at 8 - 10 AM, even though there is still energy in the storage units. This is not due to congestion issues, since some of these storage units are located at the same node as the load being shed. Rather, this is likely due to constraint \ref{eq:storage_cyclic_constrant} which specifies that the storage state of charge at the end of the day must be greater than or equal to that at the start of the day. This constraint was included so as not to overestimate the flexibility that storage could provide, though it is unlikely that such a situation would occur in reality.

@@

Considering the case with just unit commitment constraints, the reserve shedding limit can be decreased without increasing load shedding, though there is an increase (~0.09\%) in the objective function greater than the MIP gap. This is counterintuitive, as you would expect the DUC-PR model to shed all operating reserves before shedding load in the day ahead stage. The former is less costly than the latter after all, due to activation probabilities being $< 1$. However, what is occuring is most likely `economic reserve shedding' - shedding reserves because they are highly unlikely to be activated. This phenomenon was also observed in~\cite{Hermans2018}. \figref{fig:reserves_with_operating_reserves} illustrates that almost all reserve shedding occurs between 14:00 and 16:00 (hour 6830 to 6832) and not 8:00 to 10:00 as with load shedding. These are hours with a lot of solar generation and very little conventional, so to avoid this reserve shedding only requires additional generator commitment and not increasing the amount of load shedding in those hours.

@@

Multiplying the load by a factor of 1.5 gives the results shown in \tabref{tab:results_with_RANet_LM_15}. Here it is possible to discern a trade-off between load shedding and reserve shedding, albeit a small one - reducing reserve shedding to 0 from approximately 12,000 MWh increases load shedding only by 914 MWh. That this effect is slight is for similar reasons to that listed previously - reserve shedding occurs in different hours to load shedding and load shedding occurs in the same hours as before (8:00 to 11:00) and is likely due to the cyclic constraint on storage. The absolute imbalance limits also appear to have a limited impact (less than a million euro increase in objective function) on the results.

@@

\section{Comparison of \acs{DUCPR} and power flow model results} \label{sec:ULiege_comparison}

A comparison of the real time operational security as percieved by the \ac{DUCPR} model compared to those obtained from a power flow performed by the team ULiege is given in the figure below.

\begin{figure}[h]
    \centering
    \begin{subfigure}[t]{0.9\textwidth}
        \centering
        \includegraphics[width=\textwidth]{../plots/compare_ULiege_results/forecast_only_no_contingency.pdf}
        \caption{Day ahead load shedding for \ac{DUCPR} and operational security power flow model.}
    \end{subfigure}
    \begin{subfigure}[t]{0.9\textwidth}
        \centering
        \includegraphics[width=\textwidth]{../plots/compare_ULiege_results/with_scenarios_no_contingency.pdf}
        \caption{Expected load shedding for DUC-PR and operational security power flow model.}
    \end{subfigure}
\caption{Results are for day 309, with storage, unit commitment and network (reserve activation) constraints.\label{fig:Uliege_compare}}
\end{figure}

Clearly the two models disagree and by some margin. We suggest the following non-exhaustive list of reasons for why this could be:

\begin{itemize}
    \item The approximate modeling of real time imbalances in the DUC-PR model (see \secref{sec:operating_reserve_network_activation_constraints} and \secref{sec:further_constraining_operating_reserve_network_activation_constraints}). As extensively discussed, this means the DUC-PR model overestimates the ease with which real time imbalances may be dealt with.
    \item The omission of N-1 contingencies in the DUC-PR model. This omission also leads to an overestimation of the ease of dealing with real time imbalances.
    \item The extra-liberty for storage to deal with real time imbalances in ULiege's power flow, since the storage state of charge is not tracked and there are no cyclical constraints.
\end{itemize}
